---
title: "S. Popov Faecal Bacteria: EDA"
author: "Euan McDonnell"
date: "2023-07-03"
output: html_document
---


## Tasks

### To ask

* Some numbers in dad's age not arabic numerals
* What do offset date of birth mean, must be wrong with 1901.

### To Do

* Check correlations between variables, esp body size measurements and weight.
* Calculate CFUs as proportion of total CFUs per patient.


## Code

### Remove

```{r}

rm(list=ls(all.names=T)); gc()

```

### Libraries

```{r}

# Run script to get and install libraries
source("get_libraries.R")

# Get required packages that can be installed normally
lib_list <- GetLibsFromJSON(path_name="package_dependencies/", dependencies=TRUE)

```

### Parameters

```{r}

# Toggle to drop outliers idenitifed via NMDS
drop_outliers <- FALSE
mixed_is_formula <- TRUE

```

### Define directories

```{r}

# Define directories with outliers removed
scripts_dir <- paste0(getwd(), "/")
main_dir <- paste0(scripts_dir, "../")
data_dir <- paste0(main_dir, "01_data/")
res_dir <- paste0(main_dir, "03_results/", 
                  ifelse(mixed_is_formula,
                         "mixed_formula_combined/",
                         "feeding_separate/"))
eda_dir <- paste0(res_dir, "eda/")

# Add if treating formula as combined/mixed
# if (mixed_is_formula) { feed_dir <- paste0(eda_dir, "formula_combined_with_mixed/") } else { feed_dir <- eda_dir }

# Make directories
for (directory in c(scripts_dir, res_dir, eda_dir, paste0(eda_dir, "cfus/"))) { dir.create(directory, recursive = TRUE, showWarnings=FALSE) }

```

### Source extra functions

```{r}

# Source script with extra functions in
source(paste0(scripts_dir, "extra_functions.R"))

```

### Read in data

```{r}

# Read in metadata
metadata <- read_xlsx(paste0(data_dir,"Copy of AP 2023-06-20 200623 MT B.tall_format.xlsx"), sheet=1); colnames(metadata)[[1]] <- "PATIENT_ID"
metadata <- metadata[!is.na(metadata$PATIENT_ID),]

# Read in sample data
mbiome_data <- read_xlsx(paste0(data_dir,"Copy of AP 2023-06-20 200623 MT B.tall_format.xlsx"), sheet=2)
calprot_data <- read_xlsx(paste0(data_dir,"Copy of AP 2023-06-20 200623 MT B.tall_format.xlsx"), sheet=3)

```

### Fix variables

```{r}

# Fix data
metadata["mom's profesion"] <- ifelse(metadata$`mom's profesion` == "don't work", "don't work", "works")
metadata["stool_consistency"] <- gsub(",.*","",metadata$Stool)
metadata["stool_amount"] <- as.numeric(mgsub::mgsub(metadata$Stool, c(".*, ", " times.*"),c("","")))

# Toggle if treating mixed and formula as the same
if (mixed_is_formula) {
  
  # Set levels
  metadata["FEEDING"] <- "Breast"
  metadata[metadata$`Formula feeding` == "+" | metadata$`Mixt feeding` == "+",]["FEEDING"] <- "Formula"
 
# Else treat as separate
} else {
  
  # Set levels
  metadata["FEEDING"] <- "Breast"
  metadata[metadata$`Formula feeding` == "+" ,]["FEEDING"] <- "Formula"
  metadata[metadata$`Mixt feeding` == "+",]["FEEDING"] <- "Mixed"
  
}
  
# Remove pulmicort
metadata[grepl("pulm",metadata$`Treatment during pregnancy`),]["Treatment during pregnancy"] <- "multivitamins"

```

### Combine data

```{r}

# Remove space
mbiome_data$Weeks <- gsub(" ","_",mbiome_data$Weeks)
calprot_data$Weeks <- gsub(" ","_",calprot_data$Weeks)

# Add patient ID code
mbiome_data <- data.frame(SAMPLE_ID=paste(mbiome_data$Patient, mbiome_data$Weeks, sep="_"), mbiome_data)
calprot_data <- data.frame(SAMPLE_ID=paste(calprot_data$Number, calprot_data$Weeks, sep="_"), calprot_data)

# Edit awkward microbe names
colnames(mbiome_data)[4:ncol(mbiome_data)] <- gsub("\n\n","_",gsub("[.]","_",colnames(mbiome_data)[4:ncol(mbiome_data)]))

# Combine
all_data <- merge(mbiome_data, calprot_data[,c("SAMPLE_ID","calprotectine")], by="SAMPLE_ID")
all_data <- merge(metadata, all_data, by.x="PATIENT_ID", by.y="Patient")

```

### Get tall format

```{r}

# Make tall
tall.mbiome_data <- reshape2::melt(all_data[,which(colnames(all_data) %in% c(colnames(mbiome_data),"calprotectine"))], id.vars=c("SAMPLE_ID", "Weeks","calprotectine"))
colnames(tall.mbiome_data) <- c("SAMPLE_ID","WEEKS","CALPROTECTINE","ORGANISM","CFU")

```

### Update IDs etc

```{r}

# ADd dummy variable of patient ID organism
tall.mbiome_data$ORGANISM <- as.character(tall.mbiome_data$ORGANISM)
tall.mbiome_data["PATIENT_ORG"] <- paste(gsub("_.*","",tall.mbiome_data$SAMPLE_ID), tall.mbiome_data$ORGANISM, sep="_")

# Add patient ID
tall.mbiome_data["PATIENT_ID"] <- gsub("_.*","",tall.mbiome_data$SAMPLE_ID)

# Make week continuous
tall.mbiome_data["WEEKS"] <- as.numeric(gsub("_.*","",tall.mbiome_data$WEEKS))

```

### Total CFUs per patient

```{r}

# Add column totalling patient-wise CFU
group_col <- "PATIENT_ID"
tall.mbiome_data["CFU_TOTAL"] <- 0
for (id in unique(tall.mbiome_data[[group_col]])) { 
  
  # Get total patient-wise CFU
  total_cfu <- sum(tall.mbiome_data[tall.mbiome_data[[group_col]] == id,]$CFU)
  
  # View
  print(paste(id, total_cfu, sep=": "))
  
  # Add for patient 
  tall.mbiome_data[tall.mbiome_data[[group_col]] == id,]["CFU_TOTAL"] <- total_cfu
  
}

# Visualise
total_cfus <- tall.mbiome_data[!duplicated(tall.mbiome_data$PATIENT_ID),]$CFU_TOTAL
names(total_cfus) <- unique(tall.mbiome_data$PATIENT_ID)
hist(total_cfus)

```

### Define organisms of interest

```{r}

# Save dataset wiht all organisms
write.table(all_data, paste0(data_dir,"all_data.combined_mbiome_metadata.tsv"), sep="\t", quote=FALSE)

# Define organisms of interest
orgs_of_interest <- c("Total_count_of_Escherichia_coli", "conditionally_pathogenic_microflora", "lactobacteria", "bifidobacteria")

# Retain only organisms of interest
tall.mbiome_data <- tall.mbiome_data[which(tall.mbiome_data$ORGANISM %in% orgs_of_interest),]

```

#### Adjust directories to look into "outliers removed" folder

Sample IDs to be dropped are listed in a single-columned file called "samples2drop.tsv". These currently contain samples that appeared as "outliers" in the downstream NMDS, however this is subject to change and isn't the most reliable.

```{r}

# Toggle if we want to use the oulier excluded data
if (drop_outliers) {
  
  # Read in outlier file
  outlier_id.df <- read.table(paste0(data_dir,"samples2drop.tsv"), header=FALSE)
  
  # Define directories with outliers removed
  data_dir <- paste0(data_dir, "outliers_removed/")
  res_dir <- paste0(res_dir, "outliers_removed/")
  
  # Make directories
  for (directory in c(res_dir, data_dir)) { dir.create(directory, recursive = FALSE, showWarnings=FALSE) }
  
  # Drop outliers in this file
  all_data <- all_data[!all_data$SAMPLE_ID %in% outlier_id.df$V1,]
  tall.mbiome_data <- tall.mbiome_data[!tall.mbiome_data$SAMPLE_ID %in% outlier_id.df$V1,]
  
  # Drop excluded patients
  metadata <- metadata[metadata$PATIENT_ID %in% all_data$PATIENT_ID,]
  
}

```

### Birthweights by group

```{r}

# CFU vs organism
boxplot(Birthweight ~ Gender, data = all_data)
boxplot(Birthweight ~ `Treatment during pregnancy`, data = all_data)

# Tests
t.test(all_data[all_data$Gender == "m",]$Birthweight, all_data[all_data$Gender == "f",]$Birthweight)
t.test(all_data[all_data$`Treatment during pregnancy`== "-",]$Birthweight, all_data[all_data$`Treatment during pregnancy` == "multivitamins",]$Birthweight)
table(all_data$`Treatment during pregnancy`)

```

### Get undetected organisms

```{r}

# Get detected species
orgwise.sum_cfu <- colSums(mbiome_data[,4:ncol(mbiome_data)]); names(orgwise.sum_cfu) <- colnames(mbiome_data[,4:ncol(mbiome_data)])
nonzero.sum_cfu <- orgwise.sum_cfu[orgwise.sum_cfu > 0]

# Save non-zero CFU organisms
saveRDS(nonzero.sum_cfu, paste0(main_dir,"03_results/undetected_organisms.rds"))

```

### Colours

```{r}

# Get organism colours
org_colors <- hue_pal()(5)
names(org_colors) <- names(nonzero.sum_cfu)

# Day
week_colors <- c("#FEE391", "#EC7014", "#662506"); names(week_colors) <- unique(all_data$Weeks)

```

### NMDS 

Based on: 

* https://mb3is.megx.net/gustame/dissimilarity-based-methods/nmds
* https://jkzorz.github.io/2019/06/06/NMDS.html

Form of dimensional reduction. Features are:

* Is non-parameteric and works on ranks so is good for often highly skewed microbiome data.
* Takes as input a distance metric. Recommend Bray-Curtis as it accounts for presence/absence as well as abundance.
* It is non-deterministic and so produces different results each time.
 
Bray-curtis difference quantifies the difference in species populations between samples. Calculated as (where $i$, $j$ are two samples, $S_i$/$S_j$ is the total number of specimens in $i$/$j$, $C_{ij}$ is the lesser counts for each species found in both sites):
 
$$BC_{ij} = 1 - \frac{2C_{ij}}{S_i + S_j}$$ 

Stress is a measaure of goodness-of-fit for each iteration. Ideally should be less than 0.2. If it is 0, then you might have an outlier very difference from all samples. \textbf{Stress should be reported with the plot}.

```{r}

# Get matrix of species abundances
for_nmds <- as.matrix(mbiome_data[,4:ncol(mbiome_data)])
rownames(for_nmds) <- mbiome_data$SAMPLE_ID

# Plot
set.seed(777)
nmds <- metaMDS(for_nmds, distance="bray")
plot(nmds)

# Get scores as output
#extract NMDS scores (x and y coordinates) for sites from newer versions of vegan package
data.scores = as.data.frame(vegan::scores(nmds)$sites)
data.scores["SAMPLE_ID"] <- rownames(for_nmds)
data.scores <- merge(mbiome_data, data.scores, by="SAMPLE_ID")

# Plot
xx <- ggplot(data.scores, aes(x = NMDS1, y = NMDS2, label=SAMPLE_ID)) + 
             geom_label() +
             theme_bw()

ggsave(paste0(eda_dir,"nmds_plot.all_samples.jpeg"), xx, width=3000, height=3000, units="px")

```

### Drop possible outliers

```{r}

# Possible outliers
outlier_samples <- c("3_3_week", "1_2_week", "5_5_week", "1_5_week", "3_5_week", 
                     "5_3_week", "5_2_week", "3_2_week", "1_3_week")

# Plot
set.seed(777)
sub.for_nmds <- for_nmds[!rownames(for_nmds) %in% outlier_samples,]
nmds <- metaMDS(sub.for_nmds, distance="bray")
plot(nmds)

# Get scores as output
#extract NMDS scores (x and y coordinates) for sites from newer versions of vegan package
data.scores = as.data.frame(vegan::scores(nmds)$sites)
data.scores["SAMPLE_ID"] <- rownames(sub.for_nmds)
data.scores <- merge(mbiome_data, data.scores, by="SAMPLE_ID")

# Add metadata
data.scores <- merge(metadata, data.scores, by.x="PATIENT_ID", by.y="Patient")

# Get week as just numbers (but as factors)
data.scores["WEEKS"] <- as.factor(gsub("_.*","",data.scores$Weeks))

# Function
PlotNMDS <- function(in_data, point_size=8, color_by="FEEDING", shape_by="WEEKS") {
  # Plot
  plt <- ggplot(in_data[order(in_data$NMDS2),], 
               aes(x = NMDS1, y = NMDS2, fill=.data[[color_by]], shape=.data[[shape_by]])) + 
               geom_point(color="black", size=point_size, alpha=0.65) + scale_shape_manual(values=c(21,22,24)) +
               theme_bw(base_size=16) + theme(legend.position="none",
                                              axis.title=element_text(size=24))
  
  # Save
  return(plt)
  
}

sex_pal <- c("green", "darkorchid4")
delivery_cols <- c("hotpink","lightblue")
momprof_cols <- c("white","black")
diag_cols <- c("salmon","yellow")
treatpreg_cols <- c("mediumpurple","white")
stoolconsistency_cols <- c("darkred","darkorange")

# Plot
dir.create(paste0(res_dir, "nmds"), recursive=TRUE)
data.scores$PATIENT_ID <- as.factor(data.scores$PATIENT_ID)
ggsave(paste0(eda_dir, "nmds/nmds_plot.outliers_removed.colbyfeeding.jpeg"), PlotNMDS(data.scores, color_by="FEEDING"), width=3000, height=3000, units="px")
ggsave(paste0(eda_dir, "nmds/nmds_plot.outliers_removed.colbygender.jpeg"), PlotNMDS(data.scores, color_by="Gender") + scale_fill_manual(values=sex_pal), width=3000, height=3000, units="px")
ggsave(paste0(eda_dir, "nmds/nmds_plot.outliers_removed.colbydelivery.jpeg"), PlotNMDS(data.scores, color_by="Delivery") + scale_fill_manual(values=delivery_cols), width=3000, height=3000, units="px")
ggsave(paste0(eda_dir, "nmds/nmds_plot.outliers_removed.colbymomwork.jpeg"), PlotNMDS(data.scores, color_by="mom's profesion") + scale_fill_manual(values=momprof_cols), width=3000, height=3000, units="px")
ggsave(paste0(eda_dir, "nmds/nmds_plot.outliers_removed.colbydiagnosis.jpeg"), PlotNMDS(data.scores, color_by="Diagnosis") + scale_fill_manual(values=diag_cols), width=3000, height=3000, units="px")
ggsave(paste0(eda_dir, "nmds/nmds_plot.outliers_removed.colbypregtreatment.jpeg"), PlotNMDS(data.scores, color_by="Treatment during pregnancy") + scale_fill_manual(values=treatpreg_cols), width=3000, height=3000, units="px")
ggsave(paste0(eda_dir, "nmds/nmds_plot.outliers_removed.colbystoolconsistency.jpeg"), PlotNMDS(data.scores, color_by="stool_consistency") + scale_fill_manual(values=stoolconsistency_cols), width=3000, height=3000, units="px")
ggsave(paste0(eda_dir, "nmds/nmds_plot.outliers_removed.colbypatient_id.jpeg"), PlotNMDS(data.scores, color_by="PATIENT_ID"), width=3000, height=3000, units="px")

```

### PVCA

Not feasible due to low number of features.

# Sort out metadata
cov_mod.formatted <- all_data[,c("PATIENT_ID","Gender", "Delivery", "mom's profesion", "Diagnosis", "Treatment during pregnancy", "Place of life", "stool_consistency", "stool_amount")]
rownames(cov_mod.formatted) <- all_data$SAMPLE_ID
cov_mod.formatted <- new("AnnotatedDataFrame", data = cov_mod.formatted)

# Expressionset data
rownames(mbiome_data) <- mbiome_data$SAMPLE_ID
expr_unadj <- ExpressionSet(assayData = t(as.matrix(mbiome_data[rownames(cov_mod.formatted),4:ncol(mbiome_data)])), phenoData = cov_mod.formatted)

# Run PVCA
pvca_unadj <- pvcaBatchAssess(abatch = expr_unadj, batch.factors = colnames(cov_mod.formatted@data), threshold = varexpl_thr)

# Assemble to df
pvca_unadj.df <- data.frame(as.data.frame(pvca_unadj$label), t(as.data.frame(pvca_unadj$dat)), Method="Un-adjusted")
colnames(pvca_unadj.df) <- c("Effect", "Variance", "Method")

### Plot PVCA

# Join
pvca_df <- pvca_unadj.df
pvca_df["Type"] <- "Single"
pvca_df[grepl("[:]",pvca_df$Effect),]["Type"] <- "Int"

# Remove duplicate raw
pvca_df <- pvca_df[!duplicated(pvca_df[,c("Effect","Method")]),]

# Define if raw/not raw
pvca_df["Method_Group"] <- "Un-adjusted"

# Visualise
pvca_plt <- ggplot(pvca_df, aes(x=gsub("Z_","",Effect), y=Variance, fill=Method_Group)) + 
  geom_bar(stat="identity", position="dodge") + scale_fill_manual(values=c(lighten("darkblue", 0.5))) +
  facet_grid(cols=vars(Type), rows=vars(Method), space="free", scale="free") +
  xlab("") + ylab("Variance") + ylim(0,0.6) +
  theme_bw() + theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5), legend.position = "top", legend.title=element_blank())

# Save individaul
ggsave(paste0(qc_dir,"pvca.barplot.jpeg"), pvca_plt, units="px", height=2000, width=2000)

### Check detection of organisms

```{r}

# CFU vs organism
boxplot(CFU ~ ORGANISM, data = tall.mbiome_data)
lapply(unique(tall.mbiome_data$ORGANISM), function(org) {
  print(paste0(org, ": ", sum(tall.mbiome_data[tall.mbiome_data$ORGANISM == org,]$CFU)))
})

# Calprotectine vs orgnaism
boxplot(CALPROTECTINE ~ ORGANISM, data = tall.mbiome_data)
lapply(unique(tall.mbiome_data$ORGANISM), function(org) {
  print(paste0(org, ": ", sum(tall.mbiome_data[tall.mbiome_data$ORGANISM == org,]$CALPROTECTINE)))
})

# Get vector to change organism names
org2plot <- list("Bifidobacteria","Lactobacteria","Total E. coli", "Opportunistic\nPathogens")
names(org2plot) <- unique(tall.mbiome_data$ORGANISM)

# Change facet labels
changeLabels <- function(variable,value){
  return(org2plot[value])
}

# GGplot
plt <- ggplot(tall.mbiome_data, aes(x=as.character(WEEKS), y=CFU, group=WEEKS)) + 
  geom_boxplot(fill="grey") + 
  geom_beeswarm(color="black") +
  facet_grid(rows=vars(ORGANISM), labeller=changeLabels) +
  xlab("Weeks") + ylab("CFU") + 
  theme_bw(base_size=28) + theme(legend.position="none",
                                 strip.text.y=element_text(size=24, angle=0),
                                 axis.text.x=element_text(size=36),
                                 axis.title=element_text(size=42),
                                 plot.title=element_text(size=72))

# Save
ggsave(paste0(eda_dir,"distribution_of_cfus.per_organism.jpeg"), plt, width=3000, height=3000, units="px")

# Save above
saveRDS(orgwise.sum_cfu,paste0(res_dir,"sum_cfus.all_species.rds"))
saveRDS(nonzero.sum_cfu,paste0(res_dir,"sum_cfus.nonzero_species.rds"))

```

### Histograms

```{r}

hist(tall.mbiome_data$CALPROTECTINE)
for (org in colnames(mbiome_data)[4:ncol(mbiome_data)]) { print(org); hist(mbiome_data[[org]]) }

hist(all_data$calprotectine)
mean(all_data$calprotectine)
var(all_data$calprotectine)

hist(metadata$`Apgar 1`)
hist(metadata$`Apgar 5`)

```

### Organism CFU Correlations

```{r}

# Corelate organisms, CFU and calprotectine
for (org in unique(tall.mbiome_data$ORGANISM)) { 
  print(paste0(org, " : ", cor(tall.mbiome_data[tall.mbiome_data$ORGANISM == org,]$CFU, tall.mbiome_data[tall.mbiome_data$ORGANISM == org,]$CALPROTECTINE, method="spearman"))) 
}
for (org in unique(tall.mbiome_data$ORGANISM)) { 
  print(paste0(org, " : ", cor(tall.mbiome_data[tall.mbiome_data$ORGANISM == org,]$CFU, tall.mbiome_data[tall.mbiome_data$ORGANISM == org,]$WEEKS, method="spearman"))) 
}
for (org in unique(tall.mbiome_data$ORGANISM)) { 
  print(paste0(org, " : ", cor(tall.mbiome_data[tall.mbiome_data$ORGANISM == org,]$WEEKS, tall.mbiome_data[tall.mbiome_data$ORGANISM == org,]$CALPROTECTINE, method="spearman"))) 
}

# Loop over weeks
for (week in unique(all_data$Weeks)) { 
  
  # Subset by week
  week_df <- all_data[all_data$Weeks == week,]
  
  # Correlate day of start investivation and organism/calprotectine - This is important as older babies may bias towards more/less/different organisms
  org_corr <- cor(week_df$`day of start investigation`, week_df[,c("bifidobacteria","lactobacteria","Total_count_of_Escherichia_coli","conditionally_pathogenic_microflora")]) 
  calprot_corr <- cor(week_df$`day of start investigation`, week_df$calprotectine)
  
  # print
  print(paste0("### Pearson's r: (Week ",week,") Offset day vs organism : ",paste(org_corr, collapse=", "), " // for bifidobacteria, lactobacteria, Total E.Coli, conditionally pathogenic")) 
  print(paste0("### Pearson's r: (Week ",week,") Offset day vs calprotectin : ",calprot_corr))
  
}

```

### Correlate microbes CFUs pairwise, per week

```{r}

# Loop over weeks
for (week in unique(all_data$Weeks)) { 

  # Subset matrix  
  print(week)
  mbiome_m <- all_data[all_data$Weeks == week,colnames(mbiome_data)[4:ncol(mbiome_data)]]
  colnames(mbiome_m) <- gsub("\n","_",colnames(mbiome_m))

  # Keep only organisms of interest - Taken out the undetected 5
  mbiome_m <- mbiome_m[,which(colnames(mbiome_m) %in% orgs_of_interest)]

  # Get correlation matrix
  corr_m <- cor(mbiome_m, method="pearson")  
  bc_m <-  as.matrix(ecodist::bcdist(t(mbiome_m)))
  colnames(corr_m) <- colnames(mbiome_m); rownames(corr_m) <- colnames(corr_m)
  colnames(bc_m) <- colnames(mbiome_m); rownames(bc_m) <- colnames(bc_m)
  
  # Edit names
  colnames(corr_m) <- mgsub::mgsub(colnames(corr_m), c("pathogenic_","of_","_"), c("pathogenic\n","of\n"," ")); rownames(corr_m) <- colnames(corr_m)
  colnames(bc_m) <- mgsub::mgsub(colnames(bc_m), c("pathogenic_","of_","_"), c("pathogenic\n","of\n"," ")); rownames(bc_m) <- colnames(bc_m)
  
  # Plot
  diag(corr_m) <- NA; diag(bc_m) <- NA
  corr_hmap <- Heatmap(corr_m, 
                       
                # Colour scale       
                col=circlize::colorRamp2(c(1,0,-1), c(lighten("red",0.25), "white", lighten("blue",0.25))),
                
                # Define cell contents
                cell_fun = function(j, i, x, y, width, height, fill) {
                  
                  grid.text(sprintf("%.2f", corr_m[i, j]), x, y, gp = gpar(fontsize = 16))
                    
                },
                heatmap_legend_param = list(title="Pearson's r",
                                            direction = "horizontal",
                                            title_position = "topcenter",
                                            legend_height = unit(4, "cm")),
                
                # No clustering
                cluster_rows=FALSE, 
                cluster_columns=FALSE)
  bc_hmap <- Heatmap(bc_m, 
                       
                # Colour scale       
                col=circlize::colorRamp2(c(0,1), c("white", lighten("blue",0.25))),
                
                # Define cell contents
                cell_fun = function(j, i, x, y, width, height, fill) {
                  
                  grid.text(sprintf("%.2f", bc_m[i, j]), x, y, gp = gpar(fontsize = 16))
                
                },
                heatmap_legend_param = list(title="Bray-Curtis Distance",
                                            direction = "horizontal",
                                            title_position = "topcenter",
                                            legend_height = unit(4, "cm")),
                
                # No clustering
                cluster_rows=FALSE, 
                cluster_columns=FALSE)
  
  # Plot heatmap
  png(paste0(eda_dir,week,".hmap.microorganism.pairwise_similarity.pearsons_r.jpeg"),width=750,height=750,units="px")
  draw(corr_hmap, heatmap_legend_side="top")
  dev.off()
  
  # Plot heatmap
  png(paste0(eda_dir,week,".hmap.microorganism.pairwise_dissimilarity.bray-curtis.jpeg"),width=750,height=750,units="px")
  draw(bc_hmap, heatmap_legend_side="top")
  dev.off()
}

```

### Stratified calcprotectin distribution

```{r}

# Get correct ordering of Timings
all_data$Weeks <- factor(all_data$Weeks, levels=c("2_week","3_week","5_week"))

# Boxplots
## Calprotectin
gen_plt <- NiceBoxPlot(all_data, x_val="Gender", y_val="calprotectine", facet_val_cols="Weeks", col_val="Weeks", cols=week_colors,
                       x_title="Gender", y_title="Calprotectin", max_y=600, min_y=0, facet_by_separate_plot=FALSE)
fed_plt <- NiceBoxPlot(all_data, x_val="FEEDING", y_val="calprotectine", facet_val_cols="Weeks", col_val="Weeks", cols=week_colors,
                       x_title="Feeding Method", y_title="Calprotectin", max_y=600, min_y=0, facet_by_separate_plot=FALSE)
vit_plt <- NiceBoxPlot(all_data, x_val="Treatment.during.pregnancy", y_val="calprotectine", facet_val_cols="Weeks", col_val="Weeks", cols=week_colors,
                       x_title="", y_title="Calprotectin", max_y=600, min_y=0, facet_by_separate_plot=FALSE)
del_plt <- NiceBoxPlot(all_data, x_val="Delivery", y_val="calprotectine", facet_val_cols="Weeks", col_val="Weeks", cols=week_colors,
                       x_title="", y_title="Delivery Method", max_y=600, min_y=0, facet_by_separate_plot=FALSE)


# Save
ggsave(paste0(eda_dir, "calprotectin_by_gender.boxplots.jpeg"), gen_plt, units="px", height=2000, width=4800)
ggsave(paste0(eda_dir, "calprotectin_by_feeding.boxplots.jpeg"), fed_plt, units="px", height=2000, width=4800)
ggsave(paste0(eda_dir, "calprotectin_by_multivit_treat.boxplots.jpeg"), vit_plt, units="px", height=2000, width=4800)
ggsave(paste0(eda_dir, "calprotectin_by_delivery.boxplots.jpeg"), del_plt, units="px", height=2000, width=4800)

```

### Feeding (formula different from mixed)

```{r}

# Run ANOVA
if (mixed_is_formula) {
  
  # Loop over weeks
  for (week in unique(all_data$Weeks)) {
    
    print(week)
    
    # ANOVA
    aov_res <- aov(calprotectine ~ FEEDING, data=all_data[all_data$Weeks == week,])
    print(TukeyHSD(aov_res))
  }
  
}

```

### Stratified birthweight distribution

```{r}

## Birthweight
gen_plt <- NiceBoxPlot(data.frame(metadata, DUMMY="DUMMY"), x_val="Gender", y_val="Birthweight", facet_val_cols="DUMMY", col_val="DUMMY", cols=c("DUMMY"="salmon"),
                       x_title="Gender", y_title="Birthweight", min_x=2000, max_y=4500) + ggtitle("") + stat_compare_means(comparisons = list(c(1,2)), label = "p.format", method="t.test")
fed_plt <- NiceBoxPlot(data.frame(metadata, DUMMY="DUMMY"), x_val="FEEDING", y_val="Birthweight", facet_val_cols="DUMMY", col_val="DUMMY", cols=c("DUMMY"="salmon"),
                       x_title="Feeding Method", y_title="Birthweight", min_x=2000, max_y=4500) + ggtitle("") + stat_compare_means(comparisons = list(c(1,2)), label = "p.format", method="t.test")
vit_plt <- NiceBoxPlot(data.frame(metadata, DUMMY="DUMMY"), x_val="Treatment.during.pregnancy", y_val="Birthweight", facet_val_cols="DUMMY", col_val="DUMMY", cols=c("DUMMY"="salmon"),
                       x_title="Multivitamin Supplementation\nDuring Pregnancy", y_title="Birthweight", min_x=2000, max_y=4500) + ggtitle("") + stat_compare_means(comparisons = list(c(1,2)), label = "p.format", method="t.test")
del_plt <- NiceBoxPlot(data.frame(metadata, DUMMY="DUMMY"), x_val="Delivery", y_val="Birthweight", facet_val_cols="DUMMY", col_val="DUMMY", cols=c("DUMMY"="salmon"),
                       x_title="Delivery Method", y_title="Birthweight", min_x=2000, max_y=4500) + ggtitle("") + stat_compare_means(comparisons = list(c(1,2)), label = "p.format", method="t.test")

# Save
ggsave(paste0(eda_dir, "birthweight_by_gender.boxplots.jpeg"), gen_plt, units="px", height=2000, width=1600)
ggsave(paste0(eda_dir, "birthweight_by_feeding_type.boxplots.jpeg"), fed_plt, units="px", height=2000, width=1600)
ggsave(paste0(eda_dir, "birthweight_by_multivit_treat.boxplots.jpeg"), vit_plt, units="px", height=2000, width=1600)
ggsave(paste0(eda_dir, "birthweight_by_delivery.boxplots.jpeg"), del_plt, units="px", height=2000, width=1600)

```

### Stratified apgar distribution

```{r}

# Get correct ordering of Timings
all_data$Weeks <- factor(all_data$Weeks, levels=c("2_week","3_week","5_week"))

# Boxplots
## apgar 1
gen_plt <- NiceBoxPlot(all_data[all_data$Weeks == "2_week",], x_val="Gender", y_val="Apgar.1", facet_val_cols="Weeks", col_val="Weeks", cols=week_colors,
                       x_title="Gender", y_title="APGAR 1") + ggtitle("")
fed_plt <- NiceBoxPlot(all_data[all_data$Weeks == "2_week",], x_val="FEEDING", y_val="Apgar.1", facet_val_cols="Weeks", col_val="Weeks", cols=week_colors,
                       x_title="Feeding Method", y_title="APGAR 1") + ggtitle("")
vit_plt <- NiceBoxPlot(all_data[all_data$Weeks == "2_week",], x_val="Treatment.during.pregnancy", y_val="Apgar.1", facet_val_cols="Weeks", col_val="Weeks", cols=week_colors,
                       x_title="", y_title="APGAR 1") + ggtitle("")
## Just APGAR 1 distribution
apgar.dist_plt <- NiceBoxPlot(data.frame(metadata, DUMMY="DUMMY"), x_val="DUMMY", y_val="Apgar.1", facet_val_cols="DUMMY", col_val="DUMMY",
                              x_title="", y_title="APGAR 1", cols=c("DUMMY"="darkred")) + ggtitle("") + theme(axis.text.x = element_blank())

# Save
ggsave(paste0(eda_dir, "apgar1_by_gender.boxplots.jpeg"), gen_plt, units="px", height=2000, width=1600)
ggsave(paste0(eda_dir, "apgar1_by_feeding.boxplots.jpeg"), fed_plt, units="px", height=2000, width=1600)
ggsave(paste0(eda_dir, "apgar1_by_multivit_treat.boxplots.jpeg"), vit_plt, units="px", height=2000, width=1600)
ggsave(paste0(eda_dir, "apgar1_distribution.boxplot.jpeg"), apgar.dist_plt, units="px", height=2000, width=1200)

## apgar 5
gen_plt <- NiceBoxPlot(all_data[all_data$Weeks == "2_week",], x_val="Gender", y_val="Apgar.5", facet_val_cols="Weeks", col_val="Weeks", cols=week_colors,
                       x_title="Gender", y_title="APGAR 5") + ggtitle("")
fed_plt <- NiceBoxPlot(all_data[all_data$Weeks == "2_week",], x_val="FEEDING", y_val="Apgar.5", facet_val_cols="Weeks", col_val="Weeks", cols=week_colors,
                       x_title="Feeding Method", y_title="APGAR 5")
vit_plt <- NiceBoxPlot(all_data[all_data$Weeks == "2_week",], x_val="Treatment.during.pregnancy", y_val="Apgar.5", facet_val_cols="Weeks", col_val="Weeks", cols=week_colors,
                       x_title="", y_title="APGAR 5") + ggtitle("")
## JUST APGAR 5 distribution
apgar.dist_plt <- NiceBoxPlot(data.frame(metadata, DUMMY="DUMMY"), x_val="DUMMY", y_val="Apgar.5", facet_val_cols="DUMMY", col_val="DUMMY",
                              x_title="", y_title="APGAR 5", cols=c("DUMMY"="darkred")) + ggtitle("") + theme(axis.text.x = element_blank())

# Save
ggsave(paste0(eda_dir, "apgar5_by_gender.boxplots.jpeg"), gen_plt, units="px", height=2000, width=1600)
ggsave(paste0(eda_dir, "apgar5_by_feeding.boxplots.jpeg"), fed_plt, units="px", height=2000, width=1600)
ggsave(paste0(eda_dir, "apgar5_by_multivit_treat.boxplots.jpeg"), vit_plt, units="px", height=2000, width=1600)
ggsave(paste0(eda_dir, "apgar5_distribution.boxplot.jpeg"), apgar.dist_plt, units="px", height=2000, width=1200)


```

### Histogram of apgar

```{r}

hist(metadata$`Apgar 1`, breaks=10)
hist(metadata$`Apgar 5`, breaks=10)

```

### Stratified CFUs for each organism

```{r}

# Loop over orgnaisms
for (org in orgs_of_interest) {
  
  # Plot
  gen_plt <- NiceBoxPlot(all_data, x_val="Gender", y_val=org, facet_val_cols="Weeks", col_val="Weeks", cols=week_colors,
                         x_title="Gender", y_title="CFU", max_y=(max(as.matrix(all_data[,orgs_of_interest])) + 5), min_y=0)
  fed_plt <- NiceBoxPlot(all_data, x_val="FEEDING", y_val=org, facet_val_cols="Weeks", col_val="Weeks", cols=week_colors,
                         x_title="Feeding Method", y_title="CFU", inc_sig_bracket=FALSE, inc_extra=stat_compare_means(comparisons = list(1:2, 2:3, c(1, 3))),
                         max_y=(max(as.matrix(all_data[,orgs_of_interest])) + 5), min_y=0)
  vit_plt <- NiceBoxPlot(all_data, x_val="Treatment.during.pregnancy", y_val=org, facet_val_cols="Weeks", col_val="Weeks", cols=week_colors,
                         x_title="", y_title="CFU", max_y=(max(as.matrix(all_data[,orgs_of_interest])) + 5), min_y=0)
  del_plt <- NiceBoxPlot(all_data, x_val="Delivery", y_val=org, facet_val_cols="Weeks", col_val="Weeks", cols=week_colors,
                         x_title="", y_title="CFU", max_y=(max(as.matrix(all_data[,orgs_of_interest])) + 5), min_y=0)

  # Save
  ggsave(paste0(eda_dir, "cfus/", org, "/", org, ".cfu_by_gender.boxplots.jpeg"), gen_plt, units="px", height=2000, width=4800)
  ggsave(paste0(eda_dir, "cfus/", org, "/", org, ".cfu_by_feeding.boxplots.jpeg"), fed_plt, units="px", height=2000, width=4800)
  ggsave(paste0(eda_dir, "cfus/", org, "/", org, ".cfu_by_multivit_treat.boxplots.jpeg"), vit_plt, units="px", height=2000, width=4800)
  ggsave(paste0(eda_dir, "cfus/", org, "/", org, ".cfu_by_delivery.boxplots.jpeg"), del_plt, units="px", height=2000, width=4800)

}

```

### Plot caprotectine over time

```{r}

# Plot
plt <- ggplot(tall.mbiome_data, aes(x=WEEKS, y=CALPROTECTINE, color=PATIENT_ID, fill=PATIENT_ID, group=PATIENT_ORG)) + 
              geom_line() + geom_point(color="black", shape=21) + 
              xlab("Weeks") + ylab("Calprotectin") + 
              theme_bw(base_size=36) + theme(legend.position="none",
                                             axis.title=element_text(size=42),
                                             plot.title=element_text(size=72))

# Save
ggsave(paste0(eda_dir, "calprotectine_vs_weeks.colby_patient.lineplt.jpeg"), width=3000, height=3000, units="px")

# Plot
plt <- ggplot(tall.mbiome_data, aes(x=WEEKS^2, y=CALPROTECTINE, color=PATIENT_ID, fill=PATIENT_ID, group=PATIENT_ORG)) + 
              geom_line() + geom_point(color="black", shape=21) + 
              xlab("Weeks^2") + ylab("Calprotectin") +
              theme_bw(base_size=36) + theme(legend.position="none",
                                             axis.title=element_text(size=42),
                                             plot.title=element_text(size=72))

# Save
ggsave(paste0(eda_dir, "calprotectine_vs_week_squared.colby_patient.lineplt.jpeg"), width=3000, height=2000, units="px")

```

### Plot bacteria over time/calprotectine (split by organism)

```{r}

# Plot
plt <- ggplot(tall.mbiome_data, aes(x=WEEKS, y=CFU, color=PATIENT_ID, fill=PATIENT_ID, group=PATIENT_ID)) + 
              geom_line() + geom_point(color="black", shape=21) + 
              facet_wrap( ~ ORGANISM, ncol=5) +
              scale_y_continuous(breaks=seq(0, max(tall.mbiome_data$CFU), 2)) +
              # scale_x_continuous(breaks=c(2,3,5)) +
              theme_bw(base_size=16) + theme(legend.position="none",
                                             axis.title=element_text(size=24))

# Save
ggsave(paste0(eda_dir, "organism_vs_weeks.colby_patient.split_by_organism.lineplt.jpeg"), width=5000, height=3000, units="px")

# Plot
plt <- ggplot(tall.mbiome_data, aes(x=CFU, y=CALPROTECTINE, color=PATIENT_ID, fill=PATIENT_ID, group=PATIENT_ID)) + 
              geom_line() + geom_point(color="black", shape=21) + 
              facet_wrap( ~ ORGANISM, ncol=5) +
              scale_x_continuous(breaks=seq(0, max(tall.mbiome_data$CFU), 2)) +
              # scale_x_continuous(breaks=c(2,3,5)) +
              ylab("Calprotectin") +
              theme_bw(base_size=16) + theme(legend.position="none",
                                             axis.title=element_text(size=24))

# Save
ggsave(paste0(eda_dir, "organism_vs_calprotectine.colby_patient.split_by_organism.lineplt.jpeg"), width=5000, height=2000, units="px")

```

### Plot bacteria over time/calprotectine (split by patient)

```{r}

# Plot
tall.mbiome_data$PATIENT_ID <- factor(tall.mbiome_data$PATIENT_ID, levels=unique(tall.mbiome_data$PATIENT_ID))
plt <- ggplot(tall.mbiome_data, aes(x=WEEKS, y=CFU, color=ORGANISM, fill=ORGANISM, group=ORGANISM)) + 
              geom_line() + geom_point(color="black", shape=21) + 
              scale_fill_manual(values=org_colors) + scale_color_manual(values=org_colors) +
              facet_wrap( ~ PATIENT_ID, ncol=8) +
              scale_y_continuous(breaks=seq(0, max(tall.mbiome_data$CFU), 2)) +
              # scale_x_continuous(breaks=c(2,3,5)) +
              theme_bw(base_size=16) + theme(legend.position="none",
                                             axis.title=element_text(size=24))

# Save
ggsave(paste0(eda_dir, "organism_vs_weeks.colby_organism.split_by_patient.lineplt.jpeg"), width=5000, height=5000, units="px")

# Plot
plt <- ggplot(tall.mbiome_data, aes(x=CFU, y=CALPROTECTINE, color=ORGANISM, fill=ORGANISM, group=ORGANISM)) + 
              geom_line() + geom_point(color="black", shape=21) + 
              scale_fill_manual(values=org_colors) + scale_color_manual(values=org_colors) +
              facet_wrap( ~ PATIENT_ID, ncol=8) +
              scale_x_continuous(breaks=seq(0, max(tall.mbiome_data$CFU), 2)) +
              # scale_x_continuous(breaks=c(2,3,5)) +
              theme_bw(base_size=16) + theme(legend.position="none",
                                             axis.title=element_text(size=24))

# Save
ggsave(paste0(eda_dir, "organism_vs_calprotectine.colby_organism.split_by_patient.lineplt.jpeg"), width=5000, height=5000, units="px")

```

### Save data

```{r}

# Save data
colnames(all_data) <- gsub("\n","_",colnames(all_data))
all_data$`Offset date of birth` <- as.character(all_data$`Offset date of birth`)
tall.mbiome_data$ORGANISM <- gsub("\n","_",tall.mbiome_data$ORGANISM)
tall.mbiome_data$PATIENT_ORG <- gsub("\n","_",tall.mbiome_data$PATIENT_ORG)

# Save
write.table(data.scores, paste0(res_dir,"microbiome_cfus_data.nmds_scores.tsv"),sep="\t", quote=FALSE, row.names=FALSE)
write.table(all_data, paste0(data_dir,"all_data.combined_mbiome_metadata.zero_cfu_orgs_removed.tsv"), sep="\t", quote=FALSE)
write.table(metadata, paste0(data_dir,"sample_metadata.tsv"), sep="\t", quote=FALSE)
write.table(tall.mbiome_data, paste0(data_dir,"microbiome_cfus_data.tall_format.zero_cfu_orgs_removed.tsv"), sep="\t", quote=FALSE)
table(all_data$FEEDING)

```





