---
title: "Fixed Effects Linear Modelling"
author: "Euan McDonnell"
date: "2023-08-01"
output: html_document
---


## Notes

## Code

### Reset environment

```{r}

rm(list=ls(all.names=T));

```

### Libraries

```{r}

# Run script to get and install libraries
source("get_libraries.R")

# Get required packages that can be installed normally
lib_list <- GetLibsFromJSON(path_name="package_dependencies/", dependencies=TRUE)

```

### Parameters

```{r}

# Toggle to drop outliers idenitifed via NMDS
drop_outliers <- FALSE
mixed_is_formula <- FALSE

```

### Define directories

```{r}

# Toggle if 
if (drop_outliers) {
  
  # Define directories with outliers removed
  scripts_dir <- paste0(getwd(), "/")
  main_dir <- paste0(scripts_dir, "../")
  data_dir <- paste0(main_dir, "01_data/")
  res_dir <- paste0(main_dir, "03_results/", 
                    ifelse(mixed_is_formula,
                           "mixed_formula_combined/",
                           "feeding_separate/"))
  eda_dir <- paste0(res_dir, "eda/")
  lm_dir <- paste0(res_dir,"fixed_effects_models/")
  
} else {
  
  # Directories
  scripts_dir <- paste0(getwd(), "/")
  main_dir <- paste0(scripts_dir, "../")
  data_dir <- paste0(main_dir, "01_data/")
  res_dir <- paste0(main_dir, "03_results/", 
                    ifelse(mixed_is_formula,
                           "mixed_formula_combined/",
                           "feeding_separate/"))
  eda_dir <- paste0(res_dir, "eda/")
  lm_dir <- paste0(res_dir, "fixed_effects_models/")
  
}

# Make directories
for (directory in c(scripts_dir, res_dir, eda_dir, lm_dir)) { dir.create(directory, recursive = TRUE, showWarnings=FALSE) }

```

### Source extra functions

```{r}

# Source script with extra functions in
source(paste0(scripts_dir, "extra_functions.R"))

```

#### Read in data

```{r}

# Read in data
all_data <- read.csv(paste0(data_dir,"all_data.combined_mbiome_metadata.tsv"), sep="\t", header=TRUE)
metadata <- read.csv(paste0(data_dir,"sample_metadata.tsv"), sep="\t", header=TRUE)
tall.mbiome_data <- read.table(paste0(data_dir,"microbiome_cfus_data.tall_format.zero_cfu_orgs_removed.tsv"), sep="\t", header=TRUE)

# Set patient IDs to factors
tall.mbiome_data$PATIENT_ID <- as.character(tall.mbiome_data$PATIENT_ID)
metadata$PATIENT_ID <- as.character(metadata$PATIENT_ID)
all_data$PATIENT_ID <- as.character(all_data$PATIENT_ID)

# Read in organisms detected (CFU > 0) in any sample
nonzero.sum_cfu <- readRDS(paste0(main_dir,"03_results/undetected_organisms.rds")) 

```

### Colours

```{r}

# Get organism colours
org_colors <- hue_pal()(length(unique(tall.mbiome_data$ORGANISM)))
names(org_colors) <- unique(tall.mbiome_data$ORGANISM)

# Weeks
week_colors <- c("#FEE391", "#EC7014", "#662506"); names(week_colors) <- unique(all_data$Weeks)
sig.week_colors <- c(week_colors,"NS"="darkgrey")

# Weeks (significant)

```

## Linear modelling

### Get data properly formatted

```{r}

# Merge with metadata
for_lm.data <- all_data

# Fix organism names
colnames(for_lm.data) <- gsub("\n"," ",colnames(for_lm.data))

# Get weens as just number
for_lm.data["WEEKS"] <- as.numeric(gsub("_week","",for_lm.data$Weeks))

# Z-scale data
for_lm.data["Z_STOOL_NUMBER"] <- scale(for_lm.data$stool_amount, center=TRUE, scale=TRUE)
for_lm.data["Z_BIRTHWEIGHT"] <- scale(for_lm.data$Birthweight, center=TRUE, scale=TRUE)
for_lm.data["Z_MOM_AGE"] <- scale(for_lm.data$Mother.s.age, center=TRUE, scale=TRUE)
for_lm.data["Z_GESTATION_AGE"] <- scale(for_lm.data$Gestation.age, center=TRUE, scale=TRUE)
for_lm.data["Z_ECOLI"] <- scale(for_lm.data$Total_count_of_Escherichia_coli, center=TRUE, scale=TRUE)
for_lm.data["Z_CONDI"] <- scale(for_lm.data$conditionally_pathogenic_microflora, center=TRUE, scale=TRUE)
for_lm.data["Z_LACTO"] <- scale(for_lm.data$lactobacteria, center=TRUE, scale=TRUE)
for_lm.data["Z_BIFID"] <- scale(for_lm.data$bifidobacteria, center=TRUE, scale=TRUE)
for_lm.data["Z_ECOLI_ENZ"] <- scale(for_lm.data$Escherichia_coli_with_weakly_expressed_enzyme__Ability, center=TRUE, scale=TRUE)

# Scale calprotectine per week
for_lm.data["Z_CALPROTECTINE"] <- for_lm.data$calprotectine
for_lm.data["Sqrt_calprotectine"] <- for_lm.data$calprotectine
for_lm.data["Z_SQRT_CALPROTECTINE"] <- for_lm.data$calprotectine
for (week in unique(for_lm.data$WEEKS)) {
  
  # Scale at the week level
  for_lm.data[for_lm.data$WEEKS == week,]["Z_CALPROTECTINE"] <- scale(for_lm.data[for_lm.data$WEEKS == week,]$calprotectine, center=TRUE, scale=TRUE)
  for_lm.data[for_lm.data$WEEKS == week,]["Sqrt_calprotectine"] <- sqrt(for_lm.data[for_lm.data$WEEKS == week,]$calprotectine)
  for_lm.data[for_lm.data$WEEKS == week,]["Z_SQRT_CALPROTECTINE"] <- scale(for_lm.data[for_lm.data$WEEKS == week,]$Sqrt_calprotectine, center=TRUE, scale=TRUE)
  
}

```

### Split per time point

Compare covariates to calprotectin at each level of time point (2, 3 and 5 weeks)

Very simple model of Calprotectin ~ Feeding + Weight gives sig on weight at weeks = 3. Inclusion of gender removes this. Also (excluding gender), at weeks = 5, it is close to sig.

```{r}

bonf_pval <- 0.05 / length(unique(for_lm.data$WEEKS))

# Fit linear model CALPROTECTINE ~ FEEDING + Z_BIRTHWEIGHT, splitting measurements by weeks
perweek.lm_coef.df <- GroupwiseLMs(for_lm.data[,!grepl("Birthweight|Feeding|^Calprotectine|calprotectine",colnames(for_lm.data))], group_fct1="WEEKS", conf_level=bonf_pval,
                                   predictor=c("FEEDING","Z_BIRTHWEIGHT","Gender","Delivery"), 
                                   response="Z_SQRT_CALPROTECTINE", just_coef=TRUE, width=2500, height=600, dir=lm_dir,
                                   file_suffix=".weekwise_models.patientwise_covariates.fixeff_coefficients.dwplt.png")
perweek.lm_model.df <- GroupwiseLMs(for_lm.data[,!grepl("Birthweight|Feeding|^Calprotectine|calprotectine",colnames(for_lm.data))], group_fct1="WEEKS", conf_level=bonf_pval,
                                    predictor=c("FEEDING","Z_BIRTHWEIGHT","Gender","Delivery"),
                                    response="Z_SQRT_CALPROTECTINE", just_coef=FALSE, width=2500, height=600, dir=lm_dir,
                                    file_suffix=".weekwise_models.patientwise_covariates.fixeff_coefficients.dwplt.png")

```

```{r}

bonf_pval <- 0.05 / length(unique(for_lm.data$WEEKS))

# Fit linear model CALPROTECTINE ~ FEEDING + Z_BIRTHWEIGHT, splitting measurements by weeks
perweek.lm_coef.df <- GroupwiseLMs(for_lm.data, group_fct1="WEEKS", conf_level=bonf_pval,
                                   predictor=c("FEEDING","Z_BIRTHWEIGHT","Gender","Delivery","Z_CONDI","Z_BIFID","Z_LACTO","Z_ECOLI"), 
                                   response="Z_SQRT_CALPROTECTINE", just_coef=TRUE, width=2500, height=1000, dir=lm_dir,
                                   file_suffix=".weekwise_models.inc_organisms.fixeff_coefficients.dwplt.png")
perweek.lm_model.df <- GroupwiseLMs(for_lm.data, group_fct1="WEEKS", conf_level=bonf_pval,
                                    predictor=c("FEEDING","Z_BIRTHWEIGHT","Gender","Delivery","Z_CONDI","Z_BIFID","Z_LACTO","Z_ECOLI"), 
                                    response="Z_SQRT_CALPROTECTINE", just_coef=FALSE, width=2500, height=1000, dir=lm_dir,
                                    file_suffix=".weekwise_models.inc_organisms.fixeff_coefficients.dwplt.png")

```





### Including time in model

#### Split by patient

```{r}

plt <- ggplot(tall.mbiome_data, aes(x=WEEKS, y=CFU, color=ORGANISM, fill=ORGANISM, group=ORGANISM)) + 
              geom_smooth(data=tall.mbiome_data, aes(x=WEEKS, y=CFU, group=ORGANISM), method="lm", se=FALSE) +  
              geom_point(color="black", shape=21) + 
              scale_fill_manual(values=org_colors) + scale_color_manual(values=org_colors) +
              facet_wrap( ~ PATIENT_ID, ncol=8) +
              scale_y_continuous(breaks=seq(0, max(tall.mbiome_data$CFU), 2)) +
              scale_x_continuous(breaks=c(2,3,5)) +
              theme_bw(base_size=16) + theme(legend.position="none",
                                             axis.title=element_text(size=24))

# Save
ggsave(paste0(lm_dir, "organism_vs_weeks.colby_organism.split_by_patient.regression_lineplt.png"), width=3000, height=3000, units="px")

plt <- ggplot(tall.mbiome_data, aes(x=CFU, y=CALPROTECTINE,color=ORGANISM, fill=ORGANISM, group=ORGANISM)) + 
              geom_smooth(data=tall.mbiome_data, aes(x=CFU, y=CALPROTECTINE,group=ORGANISM), method="lm", se=FALSE) +  
              geom_point(color="black", shape=21) +
              scale_fill_manual(values=org_colors) + scale_color_manual(values=org_colors) + 
              facet_wrap( ~ PATIENT_ID, ncol=8) +
              scale_y_continuous(breaks=seq(0, max(tall.mbiome_data$CFU), 2)) +
              # scale_x_continuous(breaks=c(2,3,5)) +
              theme_bw(base_size=16) + theme(legend.position="none",
                                             axis.title=element_text(size=24))

# Save
ggsave(paste0(lm_dir, "organism_vs_calprotectine.colby_organism.split_by_patient.regression_lineplt.png"), width=4000, height=2000, units="px")

```

#### Split by organism

```{r}

plt <- ggplot(tall.mbiome_data, aes(x=WEEKS, y=CFU, color=PATIENT_ID, fill=PATIENT_ID, group=PATIENT_ID)) + 
              geom_smooth(data=tall.mbiome_data, aes(x=WEEKS, y=CFU, group=PATIENT_ID), method="lm", se=FALSE) +  
              geom_point(color="black", shape=21) + 
              facet_wrap( ~ ORGANISM, ncol=5) +
              scale_y_continuous(breaks=seq(0, max(tall.mbiome_data$CFU), 2)) +
              scale_x_continuous(breaks=c(2,3,5)) +
              theme_bw(base_size=16) + theme(legend.position="none",
                                             axis.title=element_text(size=24))

# Save
ggsave(paste0(lm_dir, "organism_vs_weeks.colby_patient.split_by_organism.regression_lineplt.png"), width=2000, height=4000, units="px")

plt <- ggplot(tall.mbiome_data, aes(x=CFU, y=CALPROTECTINE, color=PATIENT_ID, fill=PATIENT_ID, group=PATIENT_ID)) + 
              geom_smooth(data=tall.mbiome_data, aes(x=CFU, y=CALPROTECTINE,group=PATIENT_ID), method="lm", se=FALSE) +  
              geom_point(color="black", shape=21) + 
              facet_wrap( ~ ORGANISM, ncol=5) +
              scale_y_continuous(breaks=seq(0, max(tall.mbiome_data$CFU), 2)) +
              # scale_x_continuous(breaks=c(2,3,5)) +
              theme_bw() + theme(legend.position="none",
                                             axis.title=element_text(size=24))


# Save
ggsave(paste0(lm_dir, "organism_vs_calprotectine.colby_patient.split_by_organism.regression_lineplt.png"), width=4000, height=3000, units="px")

```


