---
title: "Modelling"
author: "Euan McDonnell"
date: "2023-07-12"
output: html_document
---

## Mixed linear model

### Notes

#### Dispersion

Based on: https://scialert.net/fulltext/?doi=ajms.2010.1.15#:~:text=The%20negative%20binomial%20distribution%20(NB,is%20known%2C%20many%20simplifications%20result.

* Over-dispersion is when variance > mean in count data - likely best modelled by a Negative Binomial (NB) distribution. In this ($\mu$ is mean, $\sigma^2$ is variance and $\psi$ is dispersion parameter):

$$\sigma^2 = \frac{\mu + \mu^2 }{ \psi }$$

* Can determine $\psi$ via MLE, method of moments estimator (MME), maximum quasi-likelihood estimators (MQLE). MME is recomended for small sample sizes (< 20), MLE when $\mu \gt \sigma^2$. With adequate sample size (>20) and $\alpha = 1 / \psi$ is small, MQLE is preferred vs MLE. MQLE is preferred when sample $\sigma^2$ is near population $\mu$. When $\mu$ is small, $\sigma^2$ is large, MME has lower bias and may perform the best.
* Tend to determine $\psi$ by MLE, but tends to only work when $\sigma^2 \gg \mu$. If $\sigma^2 \le \mu$ or if $\sigma^2 > \mu$ (under-dispersed) then it fails as MLE estimates don't exist as the log likelihood function is asymptotic with no maximum. Similar if $\mu = 0$ as $\sigma^2 = \mu = 0$. But can instead solve for $\psi$'s reciprocal, $\alpha = \frac{1}{\psi}$ as this deals with the issue of $MLE = \inf$ when $\sigma^2 = \mu$.
* When $\sigma^2$ is only slightly larger than $\mu$, then $\psi$ can be calculated to be very large, when the true value is much smaller.

#### Model fitting

* ```{r}boundary (singular) fit: see help('isSingular')``` error refers to when the fitted mixed model has estimated parameters onn the boundary of the feasible parameter space: or that variances of one or more linear cominations of effects are (close to) zero. The estimated variance-covariance matrices are less than full rank. AKA some "dimensions" of the variance-covariance matrix are estimated to be exactly 0. Easy to detect for simple models (ie <3 paramters, such as intercept + slope) but not more complex models (>=3 parameters) as models can be singular without individual variances being close to zero. ```{r}isSingular()``` helps mitigate some of this. Uses ```{r}rePCA()``` function on var-cov matrix to isolate and inspect orthogonal variance components by looking at their stdevs and mapping these to variance terms.
* Problems this causes are:
  * Singular fits may correspond to overfitted models with poor power.
  * Numerical problems/mis-convergences may be higher for singular models.
  * May be difficult to compute profile confidence intervals
  * Inferential statistics such as Wald statistics or LRTs may be innappropriate.
* No consensus on how to deal with them. Generally recommended to:
  * Simplify models (and underlying data) as much as possible.
  * Model selection to balance predictive accuracy with overfitting/Type I error
  * "keep it maximal" - i.e. fit the most complex model consistent with the experimental design, removing only terms required to allow a non-singular fit (Barr et al. 2013), or removing further terms based on p-values or AIC
  * Use a partial Bayesian method to produce MAP estimates using regularizing priors to force estimated random-effects variance-covariance matrices away from singularity (ie `blme` package.
  * Use a full Bayesian method to regularise the model via informative priors and give estimates/credible intervals for all parameters that average over the uncertainty in random effects paramters (`MCMCglmm`, `rstanarm`, `brms`).


### Code

#### Libraries

```{r}

# Run script to get and install libraries
source("get_libraries.R")

# Get required packages that can be installed normally
lib_list <- GetLibsFromJSON(path_name="package_dependencies/", dependencies=TRUE)

```

### Parameters

```{r}

# Toggle to drop outliers idenitifed via NMDS
drop_outliers <- FALSE
mixed_is_formula <- TRUE

```

### Define directories

```{r}

# Toggle if 
if (drop_outliers) {
  
  # Define directories with outliers removed
  scripts_dir <- paste0(getwd(), "/")
  main_dir <- paste0(scripts_dir, "../")
  data_dir <- paste0(main_dir, "01_data/")
  res_dir <- paste0(main_dir, "03_results/", 
                    ifelse(mixed_is_formula,
                           "mixed_formula_combined/",
                           "feeding_separate/"))
  eda_dir <- paste0(res_dir, "eda/")
  mlm_dir <- paste0(res_dir,"mixed_effects_models/")
  
} else {
  
  # Directories
  scripts_dir <- paste0(getwd(), "/")
  main_dir <- paste0(scripts_dir, "../")
  data_dir <- paste0(main_dir, "01_data/")
  res_dir <- paste0(main_dir, "03_results/", 
                    ifelse(mixed_is_formula,
                           "mixed_formula_combined/",
                           "feeding_separate/"))
  eda_dir <- paste0(res_dir, "eda/")
  mlm_dir <- paste0(res_dir, "mixed_effects_models/")
  
}

# Make directories
for (directory in c(scripts_dir, res_dir, eda_dir, mlm_dir)) { dir.create(directory, recursive = TRUE, showWarnings=FALSE) }

```

### Source extra functions

```{r}

# Source script with extra functions in
source(paste0(scripts_dir, "extra_functions.R"))

```

#### Read in data

```{r}

# Read in data
all_data <- read.csv(paste0(data_dir,"all_data.combined_mbiome_metadata.tsv"), sep="\t", header=TRUE)
metadata <- read.csv(paste0(data_dir,"sample_metadata.tsv"), sep="\t", header=TRUE)
tall.mbiome_data <- read.table(paste0(data_dir,"microbiome_cfus_data.tall_format.zero_cfu_orgs_removed.tsv"), sep="\t", header=TRUE)

# Read in organisms detected (CFU > 0) in any sample
nonzero.sum_cfu <- readRDS(paste0(main_dir, "03_results/undetected_organisms.rds")) 
table(all_data$FEEDING)

```

#### Define organisms of interest

These are those detected with considerable counts in most samples

```{r}

# Define organisms of interest
orgs_of_interest <- c("Total_count_of_Escherichia_coli", "conditionally_pathogenic_microflora", "lactobacteria", "bifidobacteria")

```

#### Scale data

```{r}

# Scale data at sample level
all_data["Z_CALPROTECTINE"] <- scale(all_data$calprotectine, center=TRUE, scale=TRUE)

```

#### Scale data at the organism-wise level

```{r}

# Add scaled covariates
tall.mbiome_data["Z_SQRT_CALPROTECTINE"] <- tall.mbiome_data$CALPROTECTINE 
tall.mbiome_data["Z_CALPROTECTINE"] <- tall.mbiome_data$CALPROTECTINE 
tall.mbiome_data["Z_CFU"] <- tall.mbiome_data$CFU 

# Loop over organisms and scale
for (org in orgs_of_interest) {
  
  # Scale
  tall.mbiome_data[tall.mbiome_data$ORGANISM == org,]["Z_SQRT_CALPROTECTINE"] <-  scale(sqrt(tall.mbiome_data[tall.mbiome_data$ORGANISM == org,]$CALPROTECTINE), center=TRUE, scale=TRUE)
  tall.mbiome_data[tall.mbiome_data$ORGANISM == org,]["Z_CALPROTECTINE"] <-  scale(tall.mbiome_data[tall.mbiome_data$ORGANISM == org,]$CALPROTECTINE, center=TRUE, scale=TRUE)
  tall.mbiome_data[tall.mbiome_data$ORGANISM == org,]["Z_CFU"] <-  scale(tall.mbiome_data[tall.mbiome_data$ORGANISM == org,]$CFU, center=TRUE, scale=TRUE)

}

# Scale Sqrt Calprotectin
all_data["Z_SQRT_CALPROTECTINE"] <- scale(sqrt(all_data$calprotectine), center=TRUE, scale=TRUE)

```

#### Scale additional covariates

```{r}

# Scale covariates
all_data["Z_STOOL_NUMBER"] <- scale(all_data$stool_amount, center=TRUE, scale=TRUE)
all_data["Z_BIRTHWEIGHT"] <- scale(all_data$Birthweight, center=TRUE, scale=TRUE)
all_data["Z_MOM_AGE"] <- scale(all_data$Mother.s.age, center=TRUE, scale=TRUE)
all_data["Z_WEEKS"] <- scale(as.numeric(gsub("_week","",all_data$Weeks)), center=TRUE, scale=TRUE)

# Scale organisms 
all_data["Z_ECOLI"] <- scale(all_data$Total_count_of_Escherichia_coli, center=TRUE, scale=TRUE)
all_data["Z_LACTO"] <- scale(all_data$lactobacteria, center=TRUE, scale=TRUE)
all_data["Z_BIFID"] <- scale(all_data$bifidobacteria, center=TRUE, scale=TRUE)
all_data["Z_CONDI"] <- scale(all_data$conditionally_pathogenic_microflora, center=TRUE, scale=TRUE)

```

#### Make apgar categorical 

This is because it appears mostly categorical

```{r}

# Histogram
hist(all_data[!duplicated(all_data$PATIENT_ID),]$Apgar.1, breaks=10)
hist(all_data[!duplicated(all_data$PATIENT_ID),]$Apgar.5, breaks=10)

# Make categorical
all_data$Fct_Apgar_1 <- as.factor(as.character(all_data$Apgar.1))
all_data$Fct_Apgar_5 <- as.factor(as.character(all_data$Apgar.5))
all_data["Z_APGAR_1"] <- scale(all_data$Apgar.1, center=TRUE, scale=TRUE)
all_data["Z_APGAR_5"] <- scale(all_data$Apgar.5, center=TRUE, scale=TRUE)

```

#### Format data for mlm'ing

```{r}

# Get in format for mlm'ingmcx 21`12 M \
for_mlm.data <- merge(all_data[,!colnames(all_data) %in% c("PATIENT_ID","Z_CALPROTECTINE","Z_SQRT_CALPROTECTINE",unique(tall.mbiome_data$ORGANISM))], tall.mbiome_data, by="SAMPLE_ID")

# Make patient character
for_mlm.data["PATIENT_ID"] <- paste0("id_",for_mlm.data$PATIENT_ID)

```

### Colours

```{r}

# Get organism colours
org_colors <- hue_pal()(5)
names(org_colors) <- names(nonzero.sum_cfu)

# Weeks
week_colors <- c("#FEE391", "#EC7014", "#662506"); names(week_colors) <- unique(all_data$Weeks)
sig.week_colors <- c(week_colors,"NS"="darkgrey")

```

### Mixed effects models

Based on:
* PDF on mixed models in meetings and notes
* Random effect estimation and correlation parameter: https://rpubs.com/yjunechoe/correlationsLMEM

Random effects estimation (in this case with random intercept and random slope) involves the estimation of 3 parameters:

* Variation of the random intercepts.
* Variation of the random slopes.
* Correlation between intercepts and slopes.

These paramters define a *bivariate* normal distribution over subject intercept and slopes. 

Note that by estimating the 3rd paramter, correlation, we limit the degrees of freedom further meaning we risk overfitting/poor fits. So in some circumstances can do without it (see https://rpubs.com/yjunechoe/correlationsLMEM).

One recommended method for fitting models is:

* Forward-testing random effects: Add random effects one by one and compare models.
* Back-testing fixed effects: Compare models by dropping least significant terms.

Note that REML can only be used when comparing models with the same fixed-effects structures. Issue is that without REML it's difficult to fit the models.

Getting no issues with model convergence for all species except "bifidobacteria". This is with and without REML and with and without the optimx, L-BFGS-B method.

#### Patient as random intercept, test APGAR

# Define p-value
bonf_pval <- 0.05 / 2

# INitialise empty output dataframe
all.apgar.mixed.lmer <- data.frame()

# Beautify variable names
all_data <- all_data[,!grepl("calprotectine|^Calprotectine|Weeks|Birthweight|Feeding",colnames(all_data))]
colnames(all_data) <- str_to_title(tolower(mgsub(colnames(all_data),c("Z_","Z ","CTINE$","SQRT_"), c("","","CTIN","SQRT_"))))
all_data$Patient_id <- as.character(all_data$Patient_id)
 
# Loop over Apgar 1 and 5, including each in separate models
for (apgar in c("Z_APGAR_1","Z_APGAR_5")) {
  
  # For all data combined (subset to keep just bifidobacteria, just deduplicates the week/patient-level measurements)
  apgar.mixed.lmer <- lmer(as.formula(paste0("Sqrt_calprotectine ~ ",apgar," + Weeks + Gender + Birthweight + Feeding + (1|Patient_id)")), 
                           data = all_data, REML=TRUE, na.action = na.exclude)
  print(summary(apgar.mixed.lmer))
  
  # Plot QC metrics
  print(plot(apgar.mixed.lmer))
  qqnorm(resid(apgar.mixed.lmer))
  qqline(resid(apgar.mixed.lmer)) 
  
  # Plot model (fixed effects)
  fe_plt <- plot_model(apgar.mixed.lmer, type = "est", show.values = TRUE, 
                       ci.lvl=1-bonf_pval, 
                       p.threshold=bonf_pval,
                       vline.color="darkgrey") + 
    theme(axis.text=element_text(size=24),
          axis.title=element_text(size=36)) +
    theme_bw()
  
  # Plot model (random effects)
  re_plt <- plot_model(apgar.mixed.lmer, type = "re", 
                       ci.lvl=1-bonf_pval,
                       p.threshold=bonf_pval,
                       vline.color="darkgrey") + 
    theme(axis.text=element_text(size=24),
          axis.title=element_text(size=36)) +
    theme_bw()
  
  # Save random effects model
  dir.create(paste0(mlm_dir,"/by_apgar/"))
  ggsave(paste0(mlm_dir,"/by_apgar/",apgar,".fixeff_coefficients.dwplt.jpeg"), fe_plt, units="px", width=2500, height=1000)
  ggsave(paste0(mlm_dir,"/by_apgar/",apgar,".randeff_coefficients.dwplt.jpeg"), re_plt, units="px", width=2000, height=4000)
  
  # Get dataframe
  apgar.mixed.lmer.df <- data.frame(broom.mixed::tidy(apgar.mixed.lmer, conf.int=TRUE, conf.level=(1 - bonf_pval)), GROUP=apgar)
  apgar.mixed.lmer.df <- apgar.mixed.lmer.df[!is.na(apgar.mixed.lmer.df$std.error),]
  
  # Add apgar
  apgar.mixed.lmer.df["model"] <- apgar
  
  # Add to larger dataframe
  all.apgar.mixed.lmer <- rbind(all.apgar.mixed.lmer, apgar.mixed.lmer.df)
  
}
  
table(all.apgar.mixed.lmer$model)

# Define plot colours
plot_colors <- c("darkred", "NS"="darkgrey", "Week: 2"="black", "Week: 3"="black", "Week: 5"="black")
names(plot_colors)[[1]] <- paste0("p < ",round(bonf_pval,4))
  
# Sort term and facetting names
all.apgar.mixed.lmer$term <- mgsub(all.apgar.mixed.lmer$term, c("FEEDING","Z_BIRTHWEIGHT","Gender","Z_WEEKS","Z_Apgar"), c("","Birthweight","","Weeks","Apgar "))
all.apgar.mixed.lmer$model <- gsub("Z_Apgar","Apgar ",all.apgar.mixed.lmer$model)  

# Plot
DWPlot(all.apgar.mixed.lmer, pval=bonf_pval,
       plot_colors=plot_colors, filename=paste0(mlm_dir,"apgars.random_effects_model.calprot_by_patient.effect_size.dw_plots.jpeg"))

#### Patient as random intercept, all organisms included

```{r}

# Define p-value
bonf_pval <- 0.05

# Relevel delivery
all_data$Delivery <- factor(all_data$Delivery, levels=c("n","cs"))

# Run model
org.mixed.lmer <- lmer(as.formula(paste0("Z_SQRT_CALPROTECTINE ~ Z_BIRTHWEIGHT + Z_WEEKS + Gender + Delivery + FEEDING + Z_ECOLI + Z_CONDI + Z_LACTO + Z_BIFID + (1|PATIENT_ID)")), 
                       data = all_data, na.action = na.exclude)
print(summary(org.mixed.lmer))

# Plot QC metrics
print(plot(org.mixed.lmer))
qqnorm(resid(org.mixed.lmer))
qqline(resid(org.mixed.lmer)) 

# Convert axis labels
axisMappings <- c("Birthweight","Weeks","Gender [m]","Delivery [cs]","Feeding [Formula]","E. coli","Opp. Pathogens","Lactobacteria","Bifidobacteria")
names(axisMappings) <- colnames(coef(org.mixed.lmer)$PATIENT_ID)[-1]
  
# Plot model (fixed effects)
fe_plt <- plot_model(org.mixed.lmer, type = "est", show.values = TRUE, 
                     ci.lvl=1-bonf_pval, 
                     p.threshold=bonf_pval,
                     show.p=TRUE,
                     value.size=7,
                     value.offset=0.3,
                     dot.size=1,
                     vline.color="darkgrey") + 
    ggtitle("") +
    scale_x_discrete(labels=axisMappings) + 
    theme(axis.text.y=element_text(size=48),
          axis.title=element_text(size=72)) +
    theme_bw()
  
# Plot model (random effects)
re_plt <- plot_model(org.mixed.lmer, type = "re", 
                     ci.lvl=1-bonf_pval,
                     show.p=TRUE,
                     value.size=4.5,
                     dot.size=1,
                     value.offset = 0.3,
                     vline.color="darkgrey") + 
    ggtitle("") +
    theme_bw()

fe_plt$theme$axis.text$size <- 20
fe_plt$theme$axis.text.y$size <- 24
fe_plt$theme$axis.title.x$size <- 36

# Save random effects model
dir.create(paste0(mlm_dir,"/all_organisms_in_model/"))
ggsave(paste0(mlm_dir,"/all_organisms_in_model/all_organisms_in_model.fixeff_coefficients.dwplt.jpeg"), fe_plt, units="px", width=3000, height=2000)
ggsave(paste0(mlm_dir,"/all_organisms_in_model/all_organisms_in_model.randeff_coefficients.dwplt.jpeg"), re_plt, units="px", width=2000, height=4000)
 
# Get dataframe
org.mixed.lmer.df <- data.frame(broom.mixed::tidy(org.mixed.lmer, conf.int=TRUE, conf.level=(1 - bonf_pval)), GROUP="")
org.mixed.lmer.df <- org.mixed.lmer.df[!is.na(org.mixed.lmer.df$std.error),]
org.mixed.lmer.df <- data.frame(model="All\nOrganisms",org.mixed.lmer.df)

# Save table
org.mixed.lmer.df$model <- gsub("\n"," ",org.mixed.lmer.df$model)
write.table(rename(org.mixed.lmer.df[,c("model","term","estimate","std.error","statistic","df","p.value","conf.low","conf.high")],
                   c("Organism"="model")),
            paste0(mlm_dir,"/all_organisms_in_model/calprotectineAsResponse_allOrganismsIncluded_lmmModelResults.tsv"), quote=FALSE, sep="\t", row.names=FALSE)

```

#### Patient as random variable, per CFU

Fit only random intercept (as random slopes tend to not result in convergence).

# Define bonferonni-adjusted p-value
bonf_pval <- (0.05/length(orgs_of_interest))

# Defined desired model formula
model_formula <- "Z_SQRT_CALPROTECTINE ~ Z_CFU + Z_WEEKS + Gender + Delivery + Z_BIRTHWEIGHT + FEEDING + (1|PATIENT_ID)"
for_mlm.data["Weeks"] <- as.numeric(gsub("_week","",for_mlm.data$Weeks))

# Define list of all models
all.org_models <- list()
all.org_models.df <- data.frame()

# Loop over and fit models per organism
for (org in orgs_of_interest) {

  # For all data combined
  print(paste0("########## ORGANISM: ",org))
  mixed.lmer <- lmer(as.formula(model_formula), data = for_mlm.data[for_mlm.data$ORGANISM == org,], REML=TRUE, na.action = na.exclude)
  print(summary(mixed.lmer))
  
  # Plot QC metrics
  print(plot(mixed.lmer))
  qqnorm(resid(mixed.lmer))
  qqline(resid(mixed.lmer)) 
  
  # Plot model (fixed effects)
  fe_plt <- plot_model(mixed.lmer, type = "est", show.values = TRUE, 
                       ci.lvl=1-bonf_pval, 
                       p.threshold=bonf_pval,
                       vline.color="darkgrey") + 
    theme(axis.text=element_text(size=24),
          axis.title=element_text(size=36)) +
    theme_bw()
  
  # Plot model (random effects)
  re_plt <- plot_model(mixed.lmer, type = "re", 
                       ci.lvl=1-bonf_pval,
                       p.threshold=bonf_pval,
                       vline.color="darkgrey") + 
    theme(axis.text=element_text(size=24),
          axis.title=element_text(size=36)) +
    theme_bw()
  
  # Save random effects model
  dir.create(paste0(mlm_dir,"/by_organism/"))
  ggsave(paste0(mlm_dir,"/by_organism/",org,".fixeff_coefficients.dwplt.jpeg"), fe_plt, units="px", width=2500, height=1000)
  ggsave(paste0(mlm_dir,"/by_organism/",org,".randeff_coefficients.dwplt.jpeg"), re_plt, units="px", width=2000, height=4000)
  
  # Add to list
  all.org_models <- list.append(all.org_models, mixed.lmer)
  all.org_models.df <- rbind(all.org_models.df, data.frame(model=org, broom.mixed::tidy(mixed.lmer, conf.int = TRUE, conf.level=1 - bonf_pval)))
  
}

# Add species as model names
names(all.org_models) <- orgs_of_interest

# Remove NA
all.org_models.df <- all.org_models.df[!is.na(all.org_models.df$std.error),]

# Define plot colours
plot_colors <- c("darkred", "NS"="darkgrey", "Week: 2"="black", "Week: 3"="black", "Week: 5"="black")
names(plot_colors)[[1]] <- paste0("p < ",round(bonf_pval,4))
  
# Sort term and facetting names
all.org_models.df$term <- mgsub(all.org_models.df$term, c("FEEDING","Z_BIRTHWEIGHT","Gender","Z_WEEKS","Z_Apgar","Z_CFU"), c("","Birthweight","","Weeks","Apgar ","CFU"))
all.org_models.df$model <- mgsub(all.org_models.df$model, c("bifidobacteria","lactobacteria","Total_count_of_Escherichia_coli","conditionally_pathogenic_microflora"), 
                                                         c("Bifidobacteria","Lactobacteria","Total E. coli", "Conditionally\npathogenic\nmicroflora"))

#### Patient as random variable, CFU as output

Fit only random intercept (as random slopes tend to not result in convergence).

```{r}

# Define bonferonni-adjusted p-value
bonf_pval <- (0.05/length(orgs_of_interest))

# Define list of all models
all.org_models <- list()
all.org_models.df <- data.frame()

# Plot titles
plotTitles <- c("B","C","D","E")
names(plotTitles) <- orgs_of_interest

# Define combined plot
multiPltList <- list()

# Loop over and fit models per organism
for (org in orgs_of_interest) {
  
  # Subset organism
  orgData <- for_mlm.data[for_mlm.data$ORGANISM == org,!grepl("Z_CFU|Z_WEEKS|Z_BIRTHWEIGHT",colnames(for_mlm.data))]

  # Relevel delivery
  orgData$Delivery <- factor(orgData$Delivery, levels=c("n","cs"))

  # Scale
  orgData["CFU"] <- scale(orgData$CFU, center=TRUE, scale=TRUE)
  orgData["Weeks"] <- scale(orgData$WEEKS, center=TRUE, scale=TRUE)
  orgData["Birthweight"] <- scale(orgData$Birthweight, center=TRUE, scale=TRUE)
  
  # Fix variable names
  colnames(orgData) <- gsub("FEEDING","Feeding",colnames(orgData))
  
  # Defined desired model formula
  model_formula <- "CFU ~ Weeks + Gender + Delivery + Birthweight + Feeding + (1|PATIENT_ID)"
  for_mlm.data["Weeks"] <- as.numeric(gsub("_week","",for_mlm.data$Weeks))

  # For all data combined
  print(paste0("########## ORGANISM: ",org))
  mixed.lmer <- lmer(as.formula(model_formula), data = orgData, REML=TRUE, na.action = na.exclude)
  print(summary(mixed.lmer))
  
  # Plot QC metrics
  print(plot(mixed.lmer))
  qqnorm(resid(mixed.lmer))
  qqline(resid(mixed.lmer))
  
  # Tidy opportunistic pathogens
  if (grepl("conditional",org)) { org <- "Opp. Pathogen" }
  if (grepl("Total",org)) { org <- "Total E. coli" }
  
  # Plot model (fixed effects)
  fe_plt <- plot_model(mixed.lmer, type = "est",
                       ci.lvl=1-bonf_pval, 
                       show.values = TRUE,
                       value.size = 6,
                       value.offset=0.3,
                       p.threshold=bonf_pval,
                       vline.color="darkgrey") + 
    theme(axis.text=element_text(size=36),
          axis.title=element_text(size=42)) +
    ggtitle(gsub("_"," ",str_to_title(org))) +
    theme_bw()
  
  # Plot model (random effects)
  re_plt <- plot_model(mixed.lmer, type = "re", 
                       ci.lvl=1-bonf_pval,
                       show.values = TRUE,
                       value.size = 4,
                       p.threshold=bonf_pval,
                       vline.color="darkgrey") + 
    theme(axis.text=element_text(size=36),
          axis.title=element_text(size=42)) +
    ggtitle(gsub("_"," ",str_to_title(org))) +
    theme_bw()

  fe_plt$theme$axis.text$size <- 18
  fe_plt$theme$axis.title.x$size <- 20
  fe_plt$theme$plot.title$size <- 24
  
  # Save random effects model
  dir.create(paste0(mlm_dir,"/by_organism/organismAsResponse/"))
  ggsave(paste0(mlm_dir,"/by_organism/organismAsResponse/",org,"_regressedOntoCFU.fixeff_coefficients.dwplt.jpeg"), fe_plt, units="px", width=1500, height=1250)
  ggsave(paste0(mlm_dir,"/by_organism/organismAsResponse/",org,"_regressedOntoCFU.randeff_coefficients.dwplt.jpeg"), re_plt, units="px", width=2000, height=4000)
  
  # Add to list
  all.org_models <- list.append(all.org_models, mixed.lmer)
  all.org_models.df <- rbind(all.org_models.df, data.frame(model=org, broom.mixed::tidy(mixed.lmer, conf.int = TRUE, conf.level=1 - bonf_pval)))
  
  # add to multi plot
  multiPltList <- rlist::list.append(multiPltList, fe_plt)
  
}



# Add species as model names
names(all.org_models) <- orgs_of_interest

# Remove NA
all.org_models.df <- all.org_models.df[!is.na(all.org_models.df$std.error),]

# Define plot colours
plot_colors <- c("darkred", "NS"="darkgrey", "Week: 2"="black", "Week: 3"="black", "Week: 5"="black")
names(plot_colors)[[1]] <- paste0("p < ",round(bonf_pval,4))
  
# Sort term and facetting names
all.org_models.df$term <- mgsub(all.org_models.df$term, c("FEEDING","Z_BIRTHWEIGHT","Gender","Z_WEEKS","Z_Apgar","Z_CFU"), c("","Birthweight","","Weeks","Apgar ","CFU"))
all.org_models.df$model <- mgsub(all.org_models.df$model, c("bifidobacteria","lactobacteria","Total_count_of_Escherichia_coli","conditionally_pathogenic_microflora"), 
                                                         c("Bifidobacteria","Lactobacteria","Total E. coli", "Conditionally\npathogenic\nmicroflora"))

# Save combined plot
ggsave(paste0(mlm_dir,"/by_organism/organismAsResponse/allOrganisms_regressedOntoCFU.randeff_coefficients.dwplt.jpeg"),
       patchwork::wrap_plots(multiPltList, nrow=2, ncol=2), width=3000, height=2500, units="px")

# Save table
write.table(rename(all.org_models.df[,c("model","term","estimate","std.error","statistic","df","p.value","conf.low","conf.high")],
            c("Organism"="model")),
            paste0(mlm_dir,"/by_organism/organismAsResponse/perOrganismCFU_asResponse_lmmModelResults.tsv"), quote=FALSE, sep="\t", row.names=FALSE)

```

#### Plot trends

Most code for plotting adapted from: https://ourcodingclub.github.io/tutorials/mixed-models/

##### Global (fixed) trends

```{r}

# Plot calprotectin (non-log)
for (org in orgs_of_interest) {
  
  print(org)
  
  # Get predicted data and plot
  pred_df <- ggpredict(all.org_models[[org]], terms="Z_CFU")
  pred_df$x <- ( pred_df$x * sd(for_mlm.data$CFU, na.rm=TRUE) ) + mean(for_mlm.data$CFU, na.rm=TRUE)
  pred_df$predicted <- ( pred_df$predicted * sd(for_mlm.data$CALPROTECTINE, na.rm=TRUE) ) + mean(for_mlm.data$CALPROTECTINE, na.rm=TRUE)
  
  # Plot trend lines
  plt <- ggplot(pred_df) +
     geom_line(aes(x = x, y = predicted)) +          # slope
     geom_ribbon(aes(x = x, ymin = conf.low, ymax = conf.high), 
                 fill = "lightgrey", alpha = 0.5) +  # error band
     geom_point(data=for_mlm.data[for_mlm.data$ORGANISM == org,], aes(x=CFU, y=CALPROTECTINE, fill=as.factor(PATIENT_ID)), shape=21, size=4, alpha=0.65) +
     scale_colour_manual(values=hue_pal()(length(unique(for_mlm.data$PATIENT_ID)))) +
     labs(title=paste0("Global trend of Calprotectin with increased CFU:\n",gsub("_"," ",org)), x = "CFU", y="Calprotectin") +
     theme_bw(base_size=20) + theme(axis.title=element_text(size=30),
                                    plot.title= element_text(size=36),
                                    legend.position="none")
  
  # Save
  ggsave(paste0(mlm_dir,org,".global_trendlines.mixed_lm_model.calprot_vs_cfu.jpeg"), plt, units="px", width=4000, height=4000)
  
}

```

##### Random (patient-wise) trends

```{r}

# Plot calprotectin (non-log)
for (org in orgs_of_interest) {
  
  print(org)
  
  # Get predicted data and plot
  pred_df <- data.frame(ggpredict(all.org_models[[org]], terms=c("Z_CFU","PATIENT_ID"), type = "re"))
  colnames(pred_df)[[6]] <- "PATIENT_ID"
  
  # Un-scaled CFU
  pred_df$x <- ( pred_df$x * sd(for_mlm.data$CFU, na.rm=TRUE) ) + mean(for_mlm.data$CFU, na.rm=TRUE)
  pred_df$predicted <- ( pred_df$predicted * sd(for_mlm.data$CALPROTECTINE, na.rm=TRUE) ) + mean(for_mlm.data$CALPROTECTINE, na.rm=TRUE)
  pred_df$conf.high <- ( pred_df$conf.high * sd(for_mlm.data$CFU, na.rm=TRUE) ) + mean(for_mlm.data$CFU, na.rm=TRUE)
  pred_df$conf.low <- ( pred_df$conf.low * sd(for_mlm.data$CFU, na.rm=TRUE) ) + mean(for_mlm.data$CFU, na.rm=TRUE)
  
  # Plot nicely
  plt <- ggplot(data=for_mlm.data[for_mlm.data$ORGANISM == org,]) +
      geom_point(mapping=aes(x=CFU, y=CALPROTECTINE, fill=as.factor(PATIENT_ID)), shape=21, size=2, alpha=0.65) +
      facet_wrap(~ PATIENT_ID, ncol=8) +
      geom_line(data=pred_df, aes(x = x, y = predicted, color=PATIENT_ID)) +
      scale_colour_manual(values=hue_pal()(length(unique(for_mlm.data$PATIENT_ID)))) +
      labs(title=paste0("Patient-wise trend of Calprotectin with increased CFU:\n",gsub("_"," ",org)), x = "CFU", y="Calprotectin") +
      theme_bw(base_size=16) + theme(axis.title=element_text(size=24),
                                    plot.title= element_text(size=28),
                                    legend.position="none")
  
  # Save
  ggsave(paste0(mlm_dir,org,".patientwise_trendlines.mixed_lm_model.calprot_vs_cfu.jpeg"), plt, units="px", width=4000, height=3000)
  
}

```


